\documentclass[10pt,twocolumn]{article}

\usepackage[letterpaper,margin=0.70in,columnsep=0.24in]{geometry}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[font=footnotesize,labelfont=bf]{caption}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}
\usepackage{fvextra}
\setlength{\parskip}{2pt}
\setlength{\parindent}{1em}

\title{\vspace{-1.3em}\textbf{SafeRL-Drive Internal Experiment Record}\vspace{-0.5em}}
\author{Dheeraj Dhillon}
\date{22 July 2026}

\begin{document}
\twocolumn[\maketitle
\begin{abstract}
\noindent
This surrogate is the technical record behind the compact SafeRL-Drive report. It retains
failed runs, causal debugging, exact protocol, commands, persistence rules, and the
predeclared traffic-pilot decision. The public report presents the result; this document
preserves why each result can or cannot be trusted. Traffic adaptation has been
implemented but not yet trained at the time of this revision.
\end{abstract}\vspace{0.5em}]

\section{Project boundary}
The final question is whether a geometry-competent SAC policy can adapt to interactive
procedural traffic while reducing collision and retaining traffic-free generalization.
This is single-agent RL: SAC controls one ego car and MetaDrive controls every background
vehicle. Every final configuration fixes \texttt{num\_agents=1} and
\texttt{is\_multi\_agent=false}. MultiAgentMetaDrive, self-play, learned background
vehicles, image-policy training, new algorithms, architecture searches, and broad reward
sweeps are excluded. MetaDrive provides the procedural simulator and native traffic
manager~\cite{li2022metadrive}; SAC is the sole final learned controller
family~\cite{haarnoja2018sac}.

The experiment stops after three new long runs: seed-0 reference traffic adaptation,
seed-0 safety-penalty adaptation, and seed-1 confirmation of the declared winner. IDM,
ExpertPolicy, frozen-source evaluations, rendering checks, and videos do not count as
training runs.

\section{Historical audit: what failed first}
The first ambitious Phase-1 task combined a five-block road, density-0.1 traffic, weak
terminal penalties, long horizons, and unstable checkpoint selection. PPO made progress
but usually left the road. SAC learned a profitable slow or stalled behavior and later
exhibited critic and entropy instability. In one audited run, SAC's critic loss rose from
below one to millions, the learned entropy coefficient grew above 13, and the final policy
mostly braked or saturated its actions.

The audit found four measurement defects:
\begin{enumerate}
\item MetaDrive's \texttt{velocity} field is already km/h; multiplying by 3.6 inflated
legacy speed reports.
\item Validation and claimed test scenarios overlapped.
\item Mean reward selected checkpoints even when success and route completion declined.
\item Environment creation and evaluation could share MetaDrive's process-global engine.
\end{enumerate}

Corrections separated train, validation, and test seeds; selected checkpoints by success,
route completion, and failure rates; retained raw action and terminal information; and
placed validation in its own simulator subprocess. MetaDrive's default vector observation
was kept unnormalized because it is already bounded.

\section{Phase 1: infrastructure control}
The corrected Phase-1 objective was deliberately easy: demonstrate that both a discrete
PPO control and continuous SAC control could complete a straight traffic-free road using
frozen held-out policies. Training could stop only after a success-first validation gate.
IDM, PPO, and SAC each achieved 100\% success on 20 test scenarios, with zero collision
and off-road outcomes. The 20-of-20 Wilson lower bound is only 83.9\%, so this was a
pipeline gate rather than robustness evidence.

Action telemetry changed the interpretation. PPO's deterministic solution was essentially
zero steering plus full throttle: a valid but trivial open-loop solution. SAC applied
small steering corrections and varied throttle/braking. Consequently SAC became the
continuous-control basis for procedural geometry. PPO remains a historical control and
is not retrained with traffic.

\section{Phase 2: procedural geometry}
Direct SAC and curriculum SAC shared a $[256,256]$ MLP, full continuous actions, reference
MetaDrive reward, 500,000 maximum steps, no traffic, 25 validation scenarios, and 100
held-out test scenarios. Direct SAC trained on three-block maps. Curriculum learning was
used as a staged optimization intervention rather than a new algorithm
family~\cite{ozturk2021curriculum}. Curriculum SAC trained
on \texttt{C}, then \texttt{SC}, then the identical three-block target while preserving
model and replay state.

\begin{table}[t]
\centering
\caption{Final verified Phase-2 means across seeds 0 and 1.}
\begin{tabular}{lrrrr}
\toprule
Condition & Succ. & Coll. & Off-road & Route \\
\midrule
Direct & 12.0\% & 15.5\% & 83.0\% & 47.0\% \\
Curriculum & 87.0\% & 0.0\% & 9.5\% & 95.1\% \\
\bottomrule
\end{tabular}
\end{table}

Seed-level success was 24\% versus 86\% for seed 0 and 0\% versus 88\% for seed 1.
Curriculum seed 0 stopped after 150,000 total steps when its validation gate passed.
Curriculum seed 1 peaked before the final update: validation reached 92\% success at
375,000 steps and declined later, while the frozen best checkpoint still achieved 88\%
on test. Success-first checkpoint preservation was therefore essential.

\subsection{Retained failed launches}
Direct seed 2 completed 500,000 steps but achieved 0\% test success, 100\% off-road, and
19.2\% route completion. Curriculum seed 2 failed the mandatory curve stage after
100,000 steps; its best validation checkpoint had 0\% success and 41.8\% route
completion. These runs were not silently rerun and are not included in the two complete
pairs. They demonstrate initialization sensitivity and limit the geometry claim.

\subsection{Native ceilings}
On the exact Phase-2 test, IDM achieved 95\% success and ExpertPolicy 100\%. Native
controllers are feasibility ceilings, not learned-policy equivalents. Curriculum SAC
approached IDM without exceeding the simulator's native control.

\section{Historical artifact ledger}
The audit used the repository's saved summaries, per-episode CSVs, latest pointers, and
notebook outputs rather than reconstructing results from memory. The following ledger is
the evidence retained before the traffic extension:
\begin{center}\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{llrr}
\toprule
Phase & Condition & Seeds & Episodes/seed \\
\midrule
1 & IDM/PPO/SAC control & one each & 20 \\
2 & Direct SAC complete & 0, 1, 2 & 100 \\
2 & Curriculum SAC complete & 0, 1 & 100 \\
2 & Curriculum failed gate & 2 & validation only \\
2 & Native IDM/Expert & shared test & 100 \\
2 & Zero-shot traffic learned & 0, 1 & 100 \\
\bottomrule
\end{tabular}
\end{center}
``Complete'' means that the run has an immutable resolved configuration, selected model,
and compatible held-out episode record. It does not mean that the policy performed well.
Direct seed 2 is deliberately retained despite total failure. Curriculum seed 2 is
retained even though it never advanced beyond the curve gate. These artifacts prevent a
clean two-seed table from erasing the observed optimization instability.

The final comparison does not pool the Phase-1 easy-road episodes, Phase-2 procedural
episodes, or zero-shot density-0.05 episodes. Their maps, traffic definitions, selection
roles, and sometimes action interfaces differ. They are historical evidence only. The
new evaluator may reuse a prior episode file solely when its strict fingerprint encodes
the exact same environment, reward, action, split, controller, and episode count.

\section{Why traffic is the final extension}
Frozen Phase-2 policies were tested zero-shot at density 0.05:
\begin{center}
\begin{tabular}{lrrrr}
\toprule
Condition & Succ. & Coll. & Off-road & Route \\
\midrule
Direct & 3.5\% & 37.0\% & 67.5\% & 37.3\% \\
Curriculum & 69.0\% & 21.0\% & 6.0\% & 87.1\% \\
\bottomrule
\end{tabular}
\end{center}
The curriculum result shows useful initialization: geometry and lidar processing transfer
partially. It is not trained traffic competence. Seed variation was material: curriculum
seeds 0 and 1 achieved 59\% and 79\% success, with 34\% and 8\% collision. Traffic
exposure is therefore a stronger final story than another algorithm or larger network.

\section{Final experiment protocol}
\subsection{Solvability and source freeze}
IDMPolicy and ExpertPolicy first run 50 scenarios at densities 0.05 and 0.10, start seed
50000, map 3, horizon 1000, respawn traffic, and fixed traffic realization. A short
ExpertPolicy chase video is a rendering gate.

The completed geometry-curriculum checkpoints from seeds 0 and 1 are then evaluated
without retraining at densities 0, 0.05, and 0.10. Each condition uses 100 test scenarios
from seed 60000. Results are reused only when their full fingerprint matches.

\subsection{Traffic stages}
Both seed-0 pilots branch from the exact same curriculum seed-0 best checkpoint. Model
policy state transfers; the source replay buffer and optimizer state do not. A new SAC
model is constructed with the traffic configuration, actor/critic/target weights are
loaded from the source policy state, and a fresh replay buffer starts at size zero.

\begin{table}[t]
\centering
\caption{Traffic curriculum.}
\begin{tabular}{lrrl}
\toprule
Stage & Density & Max new steps & Gate \\
\midrule
A & 0.02 & 100,000 & required \\
B & 0.05 & 200,000 & descriptive \\
\bottomrule
\end{tabular}
\end{table}

Both stages use respawn traffic and randomized traffic during training. Stage A must
reach success $\geq0.70$, collision $\leq0.25$, off-road $\leq0.15$, and route
completion $\geq0.80$ on 25 fixed validation episodes. Failure is terminal for automatic
progression. No replacement pilot is launched.

\subsection{Controlled reward ablation}
\textbf{SAC-Traffic} keeps vehicle/object collision penalties at 5.
\textbf{SAC-Traffic-Safe} changes those two values to 10. Success reward remains 10;
off-road and sidewalk penalties remain 5; driving reward is 1; speed reward is 0.1;
lateral reward and custom shaping are disabled. No TTC term or safety shield is added.

\subsection{SAC configuration}
\begin{center}\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lr@{\qquad}lr}
\toprule
Setting & Value & Setting & Value \\
\midrule
Policy & MLP & Layers & 256,256 \\
Learning rate & $10^{-4}$ & Batch & 512 \\
Buffer & 300,000 & Starts & 5,000 \\
$\gamma$ & 0.99 & $\tau$ & 0.005 \\
Train frequency & 1 & Target interval & 1 \\
Entropy & 0.05 & Device & auto \\
Timeout handling & true & Compact memory & false \\
\bottomrule
\end{tabular}
\end{center}
The notebook derives $N=\min(4,\max(1,\mathrm{CPUCount}-1))$, uses one MetaDrive engine
per subprocess, and sets gradient steps to $N$. This keeps the approximate update-to-new-
transition ratio near one. The goal is simulator throughput and stable learning, not
artificial GPU saturation.

\section{Resolved traffic configuration ledger}
The authoritative file is \path{configs/sac_traffic_curriculum.yaml}. One YAML serves
both pilots; \texttt{--variant} changes only the declared reward entries. The common
training environment is:
\begin{center}\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{tabular}{ll@{\quad}ll}
\toprule
Key & Value & Key & Value \\
\midrule
map & 3 & scenarios & 200 \\
start seed & 40000 & sequential seed & true \\
spawn lane random & false & horizon & 1000 \\
truncate as terminal & false & action & continuous \\
traffic mode & respawn & random traffic & true \\
accident probability & 0 & image observation & false \\
crash vehicle done & true & crash object done & true \\
off-road done & true & agents & 1 \\
multi-agent & false & render & false \\
\bottomrule
\end{tabular}
\end{center}
Traffic density is the only stage-wise environment override: 0.02 for at most 100,000
new transitions and 0.05 for at most 200,000. Training traffic is randomized; validation
and test traffic use fixed realizations. Full steering remains available. The default
bounded MetaDrive state/lidar observation is retained, and normalization of observation
and reward is disabled.

\begin{center}\footnotesize
\begin{tabular}{lrrr}
\toprule
Split & Start seed & Scenarios & Densities \\
\midrule
Training & 40000 & 200 cycling & stage value \\
Gate validation & 50000 & 25 & stage value \\
Pilot selection & 50000 & 50 & 0, .05, .10 \\
Final test & 60000 & 100 & 0, .05, .10 \\
\bottomrule
\end{tabular}
\end{center}
All evaluation inference is deterministic. Pilot selection and final test share density
values but not episode counts or scientific roles. The selection matrix chooses the
reward variant; the held-out test matrix estimates its final behavior.

The common reward ledger is success 10, off-road 5, sidewalk collision 5, driving 1,
speed 0.1, lateral reward disabled, and an empty custom shaping dictionary. Reference
vehicle and object collision penalties are 5; safety penalties are 10. No other reward
key changes. Both conditions use the same SAC implementation and $[256,256]$ network
from Stable-Baselines3~\cite{raffin2021sb3}.

Operational frequencies are evaluation every 25,000 new transitions, checkpoint every
50,000, diagnostics every 10,000, and resource sampling every 60 seconds. Replay saving
and the SB3 progress bar are enabled. Training verbosity is zero. The config carries
\texttt{n\_envs=1} and \texttt{gradient\_steps=1} as portable defaults; the canonical
notebook resolves both together at runtime and freezes the resolved values in the run.

\subsection{Pilot selection}
Both seed-0 pilots run a 50-scenario validation matrix at densities 0, 0.05, and 0.10.
A pilot qualifies at 0.05 with success $\geq0.80$, collision and off-road $\leq0.10$,
route completion $\geq0.90$, and traffic-free success no more than ten points below the
source.

If exactly one qualifies, it wins. If both qualify, lower collision wins; within two
collision points, higher success wins; a remaining tie uses route completion. If neither
qualifies, higher success wins; within five success points, lower collision wins. The
decision and compared values are written to
\texttt{runs/traffic\_extension\_selection.json}. Only that variant is run from the seed-1
geometry checkpoint. There is no automatic seed 2.

\section{Evaluation record schema}
Final evaluation uses the same 100 scenario IDs for source SAC, adapted SAC, IDMPolicy,
and ExpertPolicy at each density. Each episode stores training seed, density, condition,
checkpoint, fingerprint, return, length, cost, speed, steering magnitude, action change,
route completion, aggregate collision, vehicle and object collision, off-road, timeout,
and success.

Raw terminal flags are non-exclusive. A plotting field imposes priority:
success, collision, off-road, timeout, other. This prevents stacked outcome figures from
exceeding 100\% while preserving simulator detail. Summaries report a Wilson 95\%
success interval. Final effects are density-0.05 success gain and collision change,
density-0 success and route retention, and adapted degradation from 0.05 to 0.10.
Two-seed means and sample standard deviations are reported, with no real-world safety
claim.

\subsection{Metric definitions and interpretation}
Success, crash, vehicle crash, object crash, off-road, and maximum-step indicators are
copied from the terminal MetaDrive information. Aggregate collision is the saved crash
flag; \texttt{crash\_vehicle} remains separate because moving-traffic contact is the
primary hazard. Collision-free is the complement of aggregate crash, not a claim of safe
driving. A timeout includes an episode ending at the configured horizon without another
higher-priority terminal outcome.

Route completion is the terminal fraction of the planned route. Mean return is the
environment reward accumulated over the episode and therefore must be interpreted under
the recorded reward variant. Mean cost is MetaDrive's accumulated cost signal. Mean speed
uses the simulator's velocity value directly in km/h; the historical erroneous factor of
3.6 is not applied. Steering magnitude averages the absolute first action coordinate.
Action change averages consecutive action-vector differences and is a rough smoothness
diagnostic, not a comfort model.

The Wilson interval is computed over episode success indicators for each condition.
Across-seed standard deviation is computed only after producing one condition summary per
training seed. These two uncertainty summaries answer different questions: episode
sampling at a fixed policy versus variation between two training lineages. Neither
supports a population-level safety statement.

The terminal-outcome priority prevents double counting. A successful terminal transition
is labeled success even if a lower-priority raw flag is also present; otherwise collision
precedes off-road, which precedes timeout, then other. The raw flags remain available so
this plotting convention can be audited or changed without rerunning the simulator.

\section{Warm start and resume semantics}
The traffic runner resolves the seed-specific
\texttt{latest\_sac\_phase2\_curriculum\_seedN.txt} pointer, normally selects
\texttt{best\_model.zip}, and records its SHA-256 digest, source evaluation fingerprint,
source run path, and source summary in \texttt{source\_lineage.json}. It validates
observation class, shape, dtype, bounds, and the same action properties before loading
weights. A mismatch fails before learning.

Fresh source replay is the default. Importing source replay requires the explicit
\texttt{--load-source-replay-buffer} switch and is not used by the canonical notebook.
After traffic learning begins, its own resumable model and traffic replay buffer are
saved at each stage boundary. Resume requires exact equality with the immutable
\texttt{resolved\_config.yaml}, including reward variant, environment count, and gradient
steps.

State records total new transitions rather than inheriting the source model's old
timestep count. Stage records also include density, traffic mode, scenario range, elapsed
wall time, and approximate environment FPS.

\subsection{Run state and artifact contract}
A new traffic launch first resolves the source pointer and model, verifies spaces, creates
a new timestamped directory, writes its immutable config and lineage, and only then
initializes learning. During a stage, periodic checkpoints are diagnostic; they are not
silently promoted over the success-first validation choice. At a stage boundary the
runner saves a resumable model and replay buffer before updating curriculum state.

The state machine has four externally meaningful outcomes:
\begin{center}\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lp{0.58\columnwidth}}
\toprule
State & Notebook behavior \\
\midrule
running/interrupted & restore model and traffic replay, then resume \\
stage complete & continue at the next declared stage \\
failed gate & preserve and stop; never bypass automatically \\
complete & reuse; never overwrite or duplicate \\
\bottomrule
\end{tabular}
\end{center}
Resume fails when the live override changes the source, variant, number of environments,
vectorization mode, gradient steps, stage budgets, or another immutable field. This
strictness is intentional: accepting a seemingly harmless runtime change could alter the
number of optimizer updates per transition and invalidate the controlled comparison.

Each completed run must contain \texttt{resolved\_config.yaml},
\texttt{run\_metadata.json}, \texttt{curriculum\_state.json},
\texttt{source\_lineage.json}, a resumable model and replay buffer, and at least one
selected or final model. Evaluation adds episode CSVs, summaries, and fingerprints under
\texttt{eval/}; detailed training and resource records stay under \texttt{logs/}. The
latest pointer is outside the run directory so consumers can resolve one canonical
lineage without scanning timestamps.

\section{Pinned chase-camera implementation}
The repository pins MetaDrive commit
\texttt{85e5dadc6c7436d324348f6e3d8f8e680c06b4db}. At that commit, retaining a
\texttt{main\_camera} sensor with \texttt{use\_render=false} activates offscreen rendering.
After reset, MetaDrive tracks the ego with the chase task. The recorder obtains CPU frames
with \texttt{env.main\_camera.perceive(to\_float=False)}.

MetaDrive otherwise filters the main camera from its sensor list when both normal
rendering and image observation are disabled. The pinned implementation therefore uses
MetaDrive's image-service switch only to retain the offscreen camera and simultaneously sets
\texttt{agent\_observation=LidarStateObservation}. SAC still receives the trained vector,
never the RGB frame. Camera settings are 1280$\times$720, distance 8.5, height 2.8,
FOV 65, smoothing and follow-lane enabled, with interface clutter hidden. No top-down
fallback occurs. Errors report offscreen mode, sensor names, expected/received shape,
MetaDrive version, and pinned commit.

The macOS smoke test exposed a second pinned-version defect:
\texttt{MainCamera} calls the window-only \texttt{requestProperties} method when it tries
to hide the cursor, but an offscreen \texttt{GraphicsBuffer} has no such method. The
recording environment leaves the unused mouse-display flag enabled, which skips that
call while keeping every visible interface panel disabled. A five-step ExpertPolicy
smoke then produced real frames and confirmed a 259-element vector policy observation.

The Colab L4 smoke exposed a separate display-backend failure: the vector environment
completed, but the display-less Linux chase process exited before producing a frame.
The recorder now selects Panda3D's EGL-backed \texttt{p3headlessgl} pipe before importing
MetaDrive whenever chase rendering runs on Linux without a \texttt{DISPLAY}. Linux
systems with a display and macOS retain the platform default. The notebook also checks
every shell command's exit code immediately, so a compile, test, vector-smoke, or
chase-smoke failure stops at its real source instead of surfacing later as a missing-file
assertion. Chase diagnostics now include the active graphics pipe and window type.

Scenario selection is deterministic from episode CSVs: the lowest matching seed is used
for first, first-success, or first-failure. Every MP4 sidecar records source, seed,
density, mode, outcome, return, completion, collision, frame count, FPS, camera settings,
vector observation shape, and fingerprint. The required set is a matched before/after
pair, a representative adapted success, an adapted failure when one exists, and an expert
demonstration.

\section{Progress, logging, and profiling}
SB3 owns the only training progress bar. The runner exposes \texttt{--progress} and
\texttt{--no-progress}; no outer tqdm wraps learning. Each stage prints its name, density,
maximum budget, and concise validation result. Evaluation uses one tqdm loop when enabled.
SB3 verbose output remains zero.

Console logging is concise INFO; files retain DEBUG context and stack traces. Startup
metadata includes command, resolved config, working directory, Git commit, Python and OS,
package versions, CPU and RAM, CUDA/GPU name and memory, selected device, environment
count, observation/action shapes, lidar beams, and policy/actor/critic parameter counts.
\texttt{logs/resource\_usage.csv} samples CPU, RAM, GPU utilization and memory when
\texttt{nvidia-smi} is present, plus environment steps per second, approximately every
60 seconds.

\section{Implementation verification record}
No long traffic adaptation was run during implementation. Static compilation succeeded
for \texttt{saferl\_drive}, \texttt{scripts}, and \texttt{tests}. The exact
\texttt{pytest -q} invocation passed all 49 tests. The suite covers traffic configuration
invariants, stage resolution, warm-start lineage, fresh-replay behavior, immutable resume
checks, density matrices, exclusive outcomes, latest pointers, fingerprints, selection,
Drive critical files, video sidecars, and notebook structure.

A local five-step vector-observation traffic smoke used one learned ego slot and native
background traffic and returned the expected 259-element observation. A separate
five-step ExpertPolicy chase smoke produced an MP4 from actual main-camera frames after
the pinned macOS workaround; its sidecar recorded density 0.05 and the same vector
observation shape. An independent 16-step SB3 toy run displayed the live
\texttt{progress\_bar=true} output, verifying both \texttt{tqdm} and \texttt{rich}.

These checks establish importability, simulator construction, traffic spawning, vector
policy input, frame production, encoding, and orchestration. They do not establish that
the three long SAC runs learn. CUDA execution, L4 throughput, multi-process MetaDrive
training over the full budgets, Stage-A qualification, pilot selection, seed-1
confirmation, and final statistics remain Colab responsibilities.

\section{Colab and Drive record}
The observed Phase-2 runtime was Google Colab with an NVIDIA L4 on 22 July 2026:
Python 3.12.13, MetaDrive 0.4.3, Stable-Baselines3 2.9.0, Gymnasium 1.3.0,
PyTorch 2.11.0/CUDA 12.8, and NumPy 2.0.2. New runs record their actual environment rather
than assuming these versions.

The live Git clone is \texttt{/content/safedrive}; persistent storage is
\texttt{/content/drive/MyDrive/SafeDrive}. Code never runs from mounted Drive. A run sync
copies one immutable run to Drive and then verifies config, metadata, curriculum state,
lineage, resumable model, replay buffer, and at least one selected/final model. Small
metadata and pointers use temporary-file replacement. TensorBoard events and intermediate
checkpoints are excluded from project-level sync. The Mac restore remains analysis-only
unless training artifacts are explicitly requested.

\section{Canonical notebook}
Only \texttt{notebooks/phase2\_colab\_driver.ipynb} remains. Its sections are:
\begin{enumerate}
\setcounter{enumi}{-1}
\item direction and experiment policy;
\item constants and paths;
\item runtime inspection;
\item Drive mount;
\item clone or fast-forward pull;
\item install and version verification;
\item Drive restore;
\item compile, tests, traffic and chase smoke gates;
\item historical Phase-1/2 summaries;
\item native traffic solvability;
\item frozen-source matrix;
\item reference seed-0 pilot;
\item safety seed-0 pilot;
\item selection;
\item selected seed-1 confirmation;
\item final matrix;
\item third-person videos;
\item plots and tables;
\item report builds;
\item final Drive sync.
\end{enumerate}
Training cells inspect latest pointers. Completed runs are reused, paused runs resume with
their saved environment count, failed gates remain terminal, and unknown states fail
closed. Main entry points appear as visible \texttt{!python -m} commands.

\subsection{Restart playbook}
After a Colab disconnect, rerun Sections 1--7 only. They recreate the live clone, install
the pinned package, mount Drive, restore runs and pointers, and repeat cheap gates. Then
inspect the next long section. Its helper reads the latest pointer and metadata: a
completed match is skipped, an interrupted match emits the exact resume command, and a
failed gate remains visible. The helper does not decide that an unknown directory is safe
to continue.

The selection step is restart-safe independently of Python variables because its full
inputs and decision are serialized to
\texttt{traffic\_extension\_selection.json}. Seed 1 reads the selected variant from that
file. Likewise, final comparison resolves runs through persisted pointers rather than
depending on objects left in notebook memory.

The live clone remains disposable. Drive is the authority for completed training
artifacts, while Git is the authority for code. Project restore does not make mounted
Drive the working directory. The final sync separately copies compact comparison files,
videos and sidecars, TeX sources, bibliography, generated macro fragments, and PDFs.

\section{Exact commands}
\begin{Verbatim}[fontsize=\scriptsize,breaklines=true,breakanywhere=true]
python -m scripts.train_curriculum \
 --config configs/sac_traffic_curriculum.yaml \
 --variant reference --seed 0 \
 --n-envs 4 --vec-env subproc --progress

python -m scripts.train_curriculum \
 --config configs/sac_traffic_curriculum.yaml \
 --variant safety --seed 0 \
 --n-envs 4 --vec-env subproc --progress

python -m scripts.evaluate --run-dir <pilot> \
 --model best --split validation --episodes 50 \
 --densities 0.0 0.05 0.10 --prefix traffic_pilot \
 validation.num_scenarios=50

python -m scripts.compare_runs \
 --traffic-extension --select-pilots

python -m scripts.train_curriculum \
 --config configs/sac_traffic_curriculum.yaml \
 --variant <selected> --seed 1 \
 --n-envs 4 --vec-env subproc --progress

python -m scripts.evaluate --run-dir <run> \
 --model best --split test --episodes 100 \
 --densities 0.0 0.05 0.10 \
 --prefix traffic_final

python -m scripts.compare_runs --traffic-extension

python -m scripts.sync_drive_runs \
 --drive-project /content/drive/MyDrive/SafeDrive \
 --to-drive --run-dir <run>
\end{Verbatim}
If Colab resolves fewer than four environments, substitute the notebook's saved value.
Resume must reuse the immutable value stored with the run.

\section{Expected output inventory}
Each adapted run contains resolved config, run metadata, source lineage, curriculum
state, stage/final/best models, resumable traffic replay, validation histories, episode
CSVs, summaries, monitor logs, resource samples, plots, and video sidecars. Project
outputs are:
\begin{Verbatim}[fontsize=\scriptsize,breaklines=true,breakanywhere=true]
runs/traffic_extension_seed_results.csv
runs/traffic_extension_comparison.csv
runs/traffic_extension_comparison.json
runs/traffic_extension_selection.json
runs/traffic_extension_outcomes.png
runs/traffic_extension_success_collision.png
runs/traffic_extension_route_completion.png
runs/traffic_extension_training_returns.png
runs/traffic_extension_retention.png
\end{Verbatim}
Generated traffic LaTeX macros are consumed by the public report. Until the long Colab
runs exist, the report shows explicit pending text rather than fabricated metrics.

\section{Traffic failure taxonomy}
Final diagnosis begins with artifact compatibility, not policy interpretation. Every
condition must have the declared map, density, respawn mode, fixed test seeds, horizon,
episode count, continuous action interface, and reward lineage. Missing or mismatched
fingerprints invalidate a comparison even when its headline numbers look plausible.

Once compatibility passes, failures are separated by terminal and behavioral signature:
\begin{itemize}
\item High vehicle collision with strong completion indicates useful progress but
insufficient interaction handling. Compare density 0.02 and 0.05 stage histories and the
reference/safety collision difference.
\item High off-road with low vehicle collision suggests geometry forgetting or unstable
avoidance. Check traffic-free retention, steering magnitude, and action change before
calling it a traffic-only problem.
\item High timeout with low collision and low completion indicates a slow or stalled
policy. Inspect throttle/brake, speed, return components, and episode length; collision
avoidance by non-driving cannot qualify.
\item High object collision should be separated from vehicle collision because it may
reflect road-boundary or static-scene interaction rather than failure to respond to
moving traffic.
\item A sharp drop from density 0.05 to 0.10 is stress degradation. It is reported, not
used to launch another curriculum stage.
\end{itemize}

Training curves are read alongside validation outcomes. Critic or return changes without
success and completion changes do not establish progress. A late final policy that is
worse than an earlier checkpoint tests whether success-first checkpoint retention worked;
the saved best model, not the last model, remains the canonical evaluation target.
Resource samples distinguish slow simulation from apparent learning stalls but are never
used to justify enlarging the network or adding GPU-only work.

Seed disagreement is itself a result. Seed 0 selects a variant under the declared rule;
seed 1 tests whether that choice survives a different geometry initialization and SAC
random seed. If seed 1 misses Stage A, it remains a failed confirmation. The analysis
must not average a failed or unevaluated run as zero, quietly substitute the final model,
or promote the other seed-0 variant after seeing test outcomes.

\section{Failure handling and interpretation}
If Stage A fails, the run is saved with \texttt{failed\_gate}; Stage B is not started.
The success-first Stage-A checkpoint is also promoted to the canonical diagnostic
\texttt{best\_model.zip}, so the exact failure policy can be evaluated after a runtime
restart without pretending that the lineage completed adaptation.
If one or both seed-0 pilots fail, the protocol records that outcome and does not create
another variant. Pilot selection uses only declared metrics. If seed-1 fails, the final
result is one completed lineage plus a failed confirmation, not a hidden one-seed average.

Success improvement with collision reduction and limited retention loss supports the
transfer hypothesis. Higher success with substantial traffic-free regression instead
demonstrates forgetting. A collision reduction achieved by stalling will appear as lower
success, timeout, and route completion and will not qualify. Negative findings remain a
valid result because the intervention and stopping rule were fixed in advance.

\section{Deferred work}
Pedestrians, weather, richer maps, pixel observations, constrained RL, safety shields,
real-data simulation, broader seeds, and alternative algorithms remain future work.
They should not be added to rescue the fixed extension. Any future study should begin
from the final immutable artifacts and state a new research question.

\bibliographystyle{plain}
\bibliography{references}
\end{document}
